Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-04-19 submission/manuscript.Rmd
germline gene section text and GO function

### Male and female germline genes are aberrantly expressed in the male _Kdm5c_-KO brain

Intriguingly, many _Kdm5c_-KO testis-enriched DEGs have functions specific to male germ cells, such as _Cytochrome C, testis (Cyct)_, a (Figure 1C), a testis-specific cytochrome c isoform found in spermatozoa mitochondria and is important for flagellar beating[@goldbergCytochromeImmunofluorescentLocalization1977; @narisawaTestisSpecificCytochromeNull2002]. We therefore wondered if the high enrichment of testis genes within the _Kdm5c_-KO brain reflected a failure to demarcate the soma and germline, as the testis contains both germ cells (e.g. spermatogonia) as well as somatic cells (e.g. Leydig cells) that support hormone production and germline functions. Gene ontology analysis of testis DEGs revealed enrichment of germline-relevant ontologies in the _Kdm5c_-KO amygdala and hippocampus, including spermatid development (GO:0007286) and sperm axoneme assembly (GO:0007288) (Figure 1A).

To validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013]. We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 1B). The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012]. We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for multiple stages of germ cell development - including spermatogonia, spermatocytes, round spermatids, and elongating spermatids, none marked somatic cells (Figure 1C). Altogether, these data suggest testis genes expressed within the _Kdm5c_-KO brain are actually male germ cell genes.

Interestingly, we also observed significant enrichment of ovary genes within the male _Kdm5c_-KO brain (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact). Although there are relatively few ovary-enriched DEGs compared to testis-DEGs (Amygdala: `r gettissval("Amygdala_DEGnumb", "Testis")` testis DEGs, `r gettissval("Amygdala_DEGnumb", "Ovary")` ovary DEGs; Hippocampus: `r gettissval("Hippocampus_DEGnumb", "Testis")` testis DEGs, `r gettissval("Hippocampus_DEGnumb", "Ovary")` ovary DEGs), they also have known germline-specific functions. For example, _Zygotic arrest 1 (Zar1)_ was recently shown to regulate oocyte translation by sequestering mRNAs[@chengMammalianOocytesStore2022] (Figure 1D).

* To assess if ovary-enriched DEGs were also germline specific, we assessed their expression in wild-type and germline-depleted (Kit^W/Wv^) female embryos, when female germ cells undergo the initial stages of development and meiosis.
	* We found XYZ.
	* Altogether, these results indicate male _Kdm5c_-KO mice express both male and female germline genes within the amygdala and hippocampus.

```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
germgenes <- read.csv("../data/processed/germGENES20.csv")

```
* One difficulty in globally characterizing the misexpression of germline genes is a comprehensive list of mouse germline-enriched genes is currently lacking.

* To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating expression in wild-type and germline-depleted (Kit^W/Wv^) male and female mouse embryos from embryonic day 12, 14, and 16. 
	* **note: need to figure out why these time points. We have P6 and adult for males but not females so it would be biased towards males if we included the later ages**

We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes. To more comprehensively assess the impact of sex on germline gene misexpression, we additionally categorized if germline-enriched genes displayed biased expression in the wild-type female (XX) or male germline. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` XX-biased, `r nrow(subset(germgenes, sexBias == "XY"))` XY-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].

>>>>

### Germline genes are aberrantly expressed in the male _Kdm5c_-KO brain
```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
germgenes <- read.csv("../data/processed/germGENES20.csv")

#testis gene ontology results
testisGO <- read.csv("../results/GO_Brain_TestisDEGs.csv")

#function to get the GO ID
getGO <- function(x, df, column){
	df_test <- df
	out <- df_test[df_test$Description == x,][column][,1]
	return(out)
}

The testis contains both germ cells (e.g. spermatogonia) and somatic cells (e.g. Leydig cells) that support hormone production and germline functions. We then wondered if the high enrichment of testis genes in the _Kdm5c_-KO brain reflected a failure to demarcate between the soma and germline. Gene ontology analysis of testis-biased DEGs revealed enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID')`, q = `r signif(getGO("spermatid development", testisGO, 'p.adjust'), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID')`, q = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust'), 3)`) (Figure 2A).

To further validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013]. We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 2B). The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012]. We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C). Together, these data suggest the _Kdm5c_-KO brain expresses germline genes.

We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice. We assessed gene expression in male and female mice at embryonic day 12, 14, and 16 embryos, as well as male postnatal day 6 and adult testes.

We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes. To more comprehensively assess the impact of sex on germline gene misexpression, we additionally categorized if germline-enriched genes displayed biased expression in the wild-type female or male germline. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
